---
title: "Is there a significant difference in cancers incidence rates between different racial and ethnic groups in a specific region?"
author: "Jainaba Jawara"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(readr)
library(purrr)
#install.packages('forcats')
library(forcats)
library(tidyverse)

#install.packages('car')

by_area_file <- read_delim(file = "USCS-1999-2020-ASCII/BYAREA.TXT", delim = "|")
View(by_area_file)

```


## preproccssing
NOTES( talk about why i dropped ~)
when doing the gender graphs look to the SEX carefully 

```{r preproccssing }
by_area_df <- by_area_file %>% select(AREA,SEX,YEAR,POPULATION,RACE,EVENT_TYPE,SITE,AGE_ADJUSTED_RATE,COUNT)
by_area_df$RACE <- gsub("Non-Hispanic", "", by_area_df$RACE)
by_area_df$COUNT <- as.numeric(gsub("[^0-9.]", "", by_area_df$COUNT))
by_area <- by_area_df %>% na.omit()
by_area <- by_area %>%
  filter(YEAR != '2016-2020')


by_area <- by_area %>% mutate(incidence_100k = (COUNT / POPULATION) * 100000)
# check crude

colnames(by_area)

```

filter this more to 'male and female'[SEX]
```{r total_incidence_by_race, echo=FALSE}
total_incidence_by_race <- by_area  %>% 
  group_by(RACE) %>% 
  filter(EVENT_TYPE == 'Incidence') %>% 
  filter(SITE == "All Cancer Sites Combined") %>% 
  filter(SEX == 'Male and Female') %>% 
  summarize(total_incidence = sum(COUNT)) %>% 
  filter(RACE != 'All Races') 


total_incidence_by_race

```

```{r total_incidence_by_race2, echo=FALSE}
incidence_100k_by_race <- by_area  %>% 
  group_by(RACE) %>% 
  filter(SEX == 'Male and Female') %>%
  summarize(incidence_per_100k = mean(incidence_100k)) %>% 
  filter(RACE != 'All Races')

incidence_100k_by_race
print(class(incidence_100k_by_race))
```

```{r mean_incidence_by_cancer_race}

colnames(by_area)
mean_incidence_by_cancer_race <- by_area %>%
  group_by(SITE,RACE)%>%
  filter(RACE != 'All Races') %>%
  filter(SEX == 'Male and Female') %>%
  filter(SITE != 'All Cancer Sites Combined') %>% 
  summarize(incidence_per_100k = mean(incidence_100k))

View(mean_incidence_by_cancer_race)
bar_plot_cancer <- ggplot(mean_incidence_by_cancer_race, aes(x = SITE, y = incidence_per_100k)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Mean Cancer Incidence Rates by Cancer Site",
       x = "Cancer Site",
       y = "Mean Incidence Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

bar_plot_cancer_vs_race <- ggplot(mean_incidence_by_cancer_race, aes(x = SITE, y = incidence_per_100k, fill = RACE)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Mean Cancer Incidence Rates by Cancer Site and Race",
       x = "Cancer Site",
       y = "Mean Incidence Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

box_plot_cancer_vs_race <- ggplot(mean_incidence_by_cancer_race, aes(x = RACE, y = incidence_per_100k)) +
  geom_boxplot() +
  labs(title = "Box Plot of Cancer Incidence Rates by Racial and Ethnic Groups per 100,000 people",
       x = "Racial and Ethnic Groups",
       y = "Cancer Incidence Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90))
bar_plot_cancer
# Display the box plot
box_plot_cancer_vs_race
bar_plot_cancer
bar_plot_cancer_vs_race

```

Total Incidence by Year 

```{r Total Incidence by Year , echo=FALSE}
total_incidence_by_year <- by_area%>% 
  group_by(YEAR,RACE) %>% 
  filter(SEX == 'Male and Female') %>%
  filter(SITE != 'All Cancer Sites Combined') %>% 
  filter(RACE != 'All Races') %>% 
  filter(EVENT_TYPE == 'Incidence') %>% 
  summarize(total_incidencek = sum(incidence_100k),.groups = 'drop') %>% 
  filter(YEAR != '2016-2020')
  
total_bar_plot_year <- ggplot(total_incidence_by_year, aes(x = YEAR, y = total_incidencek)) +
  geom_bar(stat = "identity", fill = "lightcoral") +
  labs(title = "Cancer Incidence Rates by Race from 1999 -2020  per 100,000 people",
       x = "Race",
       y = "Mean Incidence Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle =60 , vjust = 0.5, hjust=1))


total_bar_plot_year_race <- ggplot(total_incidence_by_year, aes(x = YEAR, y =total_incidencek, fill = RACE)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = " Cancer Incidence Rates by Race from 1999 -2020 per 100,000 people",
       x = "Cancer Site",
       y = "Total Incidence Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

total_bar_plot_year
total_bar_plot_year_race


```
Mean Incidence by Year 

change to boxplot( so i can see max and mix)
add"geom_jitter"
make the width smaller/narrow
make the points clear(look in to alpha)
maybe look into (geom_smooth) look into "group"



general notes
- scatter plots 

- if you want to look into orange question
  -census data
  - package data usdata?
      - Look into docs for it 
      - dataset: look into the doc
      - look in the slack for another one 
  

```{r  Mean Incidence by Year, echo=FALSE}
mean_incidence_by_year <- by_area%>% 
  group_by(YEAR,RACE) %>% 
  filter(SITE != 'All Cancer Sites Combined') %>% 
  filter(RACE != 'All Races') %>% 
  summarize(mean_incidencek = mean(incidence_100k),.groups = 'drop') %>% 
  filter(YEAR != '2016-2020')
  
mean_bar_plot_year <- ggplot(mean_incidence_by_year, aes(x = YEAR, y = mean_incidencek)) +
  geom_bar(stat = "identity", fill = "lightcoral") +
  labs(title = "Mean Cancer Incidence Rates by Race from 1999 -2020 per 100,000 people",
       x = "Year",
       y = "Mean Incidence Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle =60 , vjust = 0.5, hjust=1))


mean_bar_plot_year_race <- ggplot(mean_incidence_by_year, aes(x = YEAR, y =mean_incidencek, fill = RACE)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Mean Cancer Incidence Rates by Race from 1999 -2020",
       x = "Cancer Site",
       y = "Mean Incidence Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

mean_bar_plot_year
mean_bar_plot_year_race

```


``````{r  mean_incidence_by_area, echo=FALSE}
mean_incidence_by_area <- by_area %>% 
  group_by(AREA,RACE) %>% 
  summarize(mean_incidencek = mean(incidence_100k)) %>% 
  filter(AREA != 'United States (comparable to ICD-O-2)') %>% 
  filter(RACE != 'All Races')
mean_incidence_by_area

  
mean_bar_plot_area <- ggplot(mean_incidence_by_area, aes(x = AREA, y = mean_incidencek)) +
  geom_bar(stat = "identity", fill = "lightcoral") +
  labs(title = "Mean Cancer Incidence Rates by Race from 1999 -2020",
       x = "State",
       y = "Mean Incidence Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle =60 , vjust = 0.5, hjust=1))


mean_bar_plot_area_race <- ggplot(mean_incidence_by_area, aes(x = AREA, y =mean_incidencek, fill = RACE)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Mean Cancer Incidence Rates by State and Race ",
       x = "State",
       y = "Mean Incidence Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 0.5))
  
mean_incidence_by_year <- by_area%>% 
  group_by(YEAR,RACE,SITE) %>% 
  filter(SITE != 'All Cancer Sites Combined') %>% 
  filter(RACE != 'All Races') %>% 
  filter(SEX == "Male and Female") %>% 
  summarize(mean_incidencek = mean(incidence_100k),.groups = 'drop') %>% 
  filter(YEAR != '2016-2020')

  
mean_bar_plot_year <- ggplot(mean_incidence_by_year, aes(x = YEAR, y = mean_incidencek)) +
  geom_bar(stat = "identity", fill = "lightcoral") +
  labs(title = "Mean Cancer Incidence Rates by Race from 1999 -2020 per 100,000 people",
       x = "Race",
       y = "Mean Incidence Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle =60 , vjust = 0.5, hjust=1))
# add incidi. 
top_5_highest <- mean_incidence_by_year %>%
  filter(SITE != 'All Cancer Sites Combined') %>% 
  filter(YEAR == '2019') %>% 
  arrange(desc(mean_incidencek)) %>%
  slice(1:5)
head(top_5_highest)
  
lowest_5_lowest <- mean_incidence_by_year %>%
  filter(SITE != 'All Cancer Sites Combined') %>% 
  arrange(mean_incidencek) %>%
  slice(1:5)
View(top_5_highest)

highest_cancers <- ggplot(top_5_highest, aes(x = SITE, y = mean_incidencek,fill = RACE)) +
  geom_bar(stat = "identity", position = position_dodge2(preserve = "single")) +
  labs(title = "Top 5 Mean Cancer Incidence Rates by Race from 1999 - 2020 per 100,000 people",
       x = "Race",
       y = "Mean Incidence Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust = 1))


highest_cancers <- ggplot(top_5_highest, aes(x = SITE, y = mean_incidencek,fill = RACE)) +
  geom_bar(stat = "identity", position = position_dodge2(preserve = "single")) +
  labs(title = "Top 5 Mean Cancer Incidence Rates by Race from 1999 - 2020 per 100,000 people",
       x = "Race",
       y = "Mean Incidence Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust = 1))


lowest_cancers <- ggplot(lowest_5_lowest, aes(x = YEAR, y = mean_incidencek)) +
  geom_bar(stat = "identity", fill = "lightcoral") +
  labs(title = "Lowest 5 Mean Cancer Incidence Rates by Race from 1999 - 2020 per 100,000 people",
       x = "Race",
       y = "Mean Incidence Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust = 1))



mean_bar_plot_area
mean_bar_plot_area_race


top_5_highest_race <- ggplot(top_5_highest, aes(x = SITE, y = mean_incidencek, fill = RACE)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Mean Cancer Incidence Rates by Cancer Site and Race",
       x = "Cancer Site",
       y = "Mean Incidence Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  

lowest_5_highest_race <- ggplot(lowest_5_lowest, aes(x = SITE, y = mean_incidencek, fill = RACE)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Mean Cancer Incidence Rates by Cancer Site and Race",
       x = "Cancer Site",
       y = "Mean Incidence Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  
#highest_cancers
#lowest_cancers
#top_5_highest_race
#lowest_5_highest_race

``````

HEATMAP of location vs race 

```{r  mean_incidence_by_area_state, echo=FALSE}
mean_incidence_by_area_state <- by_area %>% 
  filter(RACE != 'All Races') %>% 
  filter(SEX == "Male and Female") %>% 
  filter(SITE == 'All Cancer Sites Combined') %>% 
  filter(AREA != 'United States (comparable to ICD-O-2)') %>% 
  filter(EVENT_TYPE == 'Incidence') %>% 
  group_by(AREA,RACE) %>%
  summarise(mean_incidence = mean(incidence_100k, na.rm = TRUE)) 


View(mean_incidence_by_area_state)


mean_heatmap_area_race <- mean_incidence_by_area_state %>% 
  ggplot(aes(x = AREA , y = RACE , fill = mean_incidence)) +
  geom_tile() +
  labs(title = "Mean Cancer Incidence Rates by State and Race",
       x = "State",
       y = "Race",
       fill = "Mean Incidence Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 0.5))

mean_heatmap_area_race

```

```{r  mean_incidence_by_area_FB, echo=FALSE}
mean_incidence_by_area_state_FB <- by_area %>% 
  filter(RACE != 'All Races') %>% 
  filter(SEX == "Female") %>% 
  filter(SITE == 'Female Breast') %>% 
  filter(AREA != 'United States (comparable to ICD-O-2)') %>% 
  filter(EVENT_TYPE == 'Incidence') %>% 
  group_by(AREA,RACE) %>%
  summarise(mean_incidence = mean(incidence_100k, na.rm = TRUE)) 


mean_heatmap_area_race_FB <- mean_incidence_by_area_state_FB %>% 
  ggplot(aes(x = AREA , y = RACE , fill = mean_incidence)) +
  geom_tile() +
  labs(title = "Mean Female Breast Cancer Incidence Rates by State and Race",
       x = "State",
       y = "Race",
       fill = "Mean Incidence Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 0.5))


mean_heatmap_area_race_FB
```

```{r  mean_incidence_by_areaFBM, echo=FALSE}
mean_incidence_by_area_state_FBM <- by_area %>% 
  filter(RACE != 'All Races') %>% 
  filter(SEX == "Female") %>% 
  filter(SITE == 'Female Breast') %>% 
  filter(AREA != 'United States (comparable to ICD-O-2)') %>% 
  filter(EVENT_TYPE == 'Mortality') %>% 
  group_by(AREA,RACE) %>%
  summarise(mean_incidence = mean(incidence_100k, na.rm = TRUE)) %>% 
    drop_na()


mean_heatmap_area_race_FBM <- mean_incidence_by_area_state_FBM %>% 
  ggplot(aes(x = AREA , y = RACE , fill = mean_incidence)) +
  geom_tile() +
  labs(title = "Mean Breast Cancer Mortality Rates by State and Race",
       x = "State",
       y = "Race",
       fill = "Mean Incidence Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 0.5))


mean_heatmap_area_race_FBM

```

```{r  ANOVA, echo=FALSE}
anova_df <- by_area %>% 
  filter(AREA != 'United States (comparable to ICD-O-2)') %>% 
  filter(RACE != 'All Races') %>% 
  filter(SEX == "Male and Female") %>% 
  filter(YEAR == '2019') %>% 
  filter(EVENT_TYPE == 'Mortality') %>% 
  filter(SITE != 'All Cancer Sites Combined')

View(anova_df)

res_aov <- aov(incidence_100k ~ RACE,
  data = anova_df
)
summary(res_aov)

hist(res_aov$residuals)

#library(car)
#qqPlot(res_aov$residuals,
  #id = FALSE # id = FALSE to remove point identification
#)

```